---
layout: topic
title: The `data.frame` class
minutes: 30
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson')
surveys <- read.csv(file="../../data/biology/surveys.csv")
```

> ## Learning Objectives
> * Indexing and sequences
> * Accessing elements of a dataframe
> * Plotting using `table`, `barplot`, `mosaic`
> * Subsetting data and reordering data


# Indexing and sequences

## Vectors

If we want to extract one or several values from a vector, we must provide one
or several indices in square brackets, just as we do in math. For instance:

```{r, results='show', purl=FALSE}
food[2]
food[c(3, 2)] #concantenate the values we want to explore
food[2:4]
```

R indexes start at 1. Programming languages like Fortran, MATLAB, and R start
counting at 1, because that's what human beings typically do. Languages in the C
family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.

`:` is a special function that creates numeric vectors of integer in increasing
or decreasing order, test `1:10` and `10:1` for instance. The function `seq()`
(for __seq__uence) can be used to create more complex patterns:

```{r, results='show', purl=FALSE}
seq(1, 10, by=2)
seq(5, 10, length.out=3)
seq(50, by=5, length.out=10)
seq(1, 8, by=3) # sequence stops to stay below upper limit
```
Another useful function in creating sequences is `rep`, which replicates elements of vector (or list, but not covered here).
In it's simplest form you can replicate a value a specified number of times. More complex operations include replicating a sequence in a specified order.

```{r, results='show', purl=FALSE}
rep(1:4, 2)
rep(1:4, each = 2)       # not the same.
rep(1:4, c(2,1,2,1))	 # specify replicate number for each element individually
rep(1:4, each = 2, len = 4)    # first 4 only
rep(1:4, each = 2, len = 10)   # 8 integers plus two recycled 1's.
rep(1:4, each = 2, times = 3)  # length 24, 3 complete replications
```

### Challenge

Using the new functions you have learned we can now **create** a `data.frame` manually with the function `data.frame()`. This function takes in each column as an argument by giving it name, and the `c()` function concatenates the values into a vector.
Below we have re-created our metadata `data.frame`.

```{r, create-data}

create_data <- data.frame(genotype=rep(c("Wt", "KO"), each=3, len=12),
			  celltype=c(rep("typeA", 6), rep("typeB", 6)),
			  replicate=rep(seq(1,3), 4),
			  row.names=paste("sample", 1:12, sep=""))
```

## Dataframes

Our metadata frame has rows and columns (it has 2 dimensions), if we want to
extract some specific data from it, we need to specify the "coordinates" we want
from it. Row numbers come first, followed by column numbers.

```{r, purl=FALSE}
metadata[1, 1]   # first element in the first column of the data frame
metadata[1, 3]   # first element in the 6th column
metadata[1:3, ] # first three rows
metadata[3, ]    # the 3rd element for all columns
metadata[ ,3]    # the entire 3rd column
```

### Challenge

1. The function `nrow()` on a `data.frame` returns the number of rows. Use it,
   in conjuction with `seq()` to create a new `data.frame` called
   `meta_by_2` that includes every other row of the metadata

```{r, echo=FALSE, purl=TRUE}
metadata[seq(1, nrow(metadata2), by=2),]

## or step-by-step
select_rows <- seq(1, nrow(metadata2), by=2)
metadata[select_rows,]
```
In particular for larger datasets, it can be tricky to remember the column
number that corresponds to a particular variable. (Are species names in column 5
or 7? oh, right... they are in column 6). In some cases, in which column the
variable will be can change if the script you are using adds or removes
columns. It's therefore often better to use column names to refer to a
particular variable, and it makes your code easier to read and your intentions
clearer.

You can do operations on a particular column, by selecting it using the `$`
sign. In this case, the entire column is a vector. For instance, to extract all
the weights from our datasets, we can use: `surveys$wgt`. You can use
`names(surveys)` or `colnames(surveys)` to remind yourself of the column names.

In some cases, you may way to select more than one column. You can do this using
the square brackets: `surveys[, c("wgt", "sex")]`.

When analyzing data, though, we often want to look at partial statistics, such
as the maximum value of a variable per species or the average value per plot.

One way to do this is to select the data we want, and create a new temporary
array, using the `subset()` function. For instance, if we just want to look at
the animals of the species "DO":

```{r, purl=FALSE}
surveys_DO <- subset(surveys, species == "DO")
```

### Challenge

1. What does the following do?

  ```{r, echo=FALSE, purl=TRUE}
  ## What does the following do? (Try to guess without executing it)
  ## surveys_DO$month[2] <- 8
  ```

