---
layout: topic
title: The `data.frame` class
minutes: 30
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson')
surveys <- read.csv(file="../../data/biology/surveys.csv")
```

> ## Learning Objectives
> * Indexing and using sequences
> * Accessing elements of a dataframe
> * Plotting using `table`, `barplot`, `mosaic`
> * Subsetting data and reordering data


# Indexing and sequences

## Vectors

If we want to extract one or several values from a vector, we must provide one
or several indices in square brackets, just as we do in math. For instance:

```{r, results='show', purl=FALSE}
food[2]
food[c(3, 2)] #concantenate the values we want to explore
food[2:4]
```

R indexes start at 1. Programming languages like Fortran, MATLAB, and R start
counting at 1, because that's what human beings typically do. Languages in the C
family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.

`:` is a special function that creates numeric vectors of integer in increasing
or decreasing order, test `1:10` and `10:1` for instance. The function `seq()`
(for __seq__uence) can be used to create more complex patterns:

```{r, results='show', purl=FALSE}
seq(1, 10, by=2)
seq(5, 10, length.out=3)
seq(50, by=5, length.out=10)
seq(1, 8, by=3) # sequence stops to stay below upper limit
```
Another useful function in creating sequences is `rep`, which replicates elements of vector (or list, but not covered here).
In it's simplest form you can replicate a value a specified number of times. More complex operations include replicating a sequence in a specified order.

```{r, results='show', purl=FALSE}
rep(1:4, 2)
rep(1:4, each = 2)       # not the same.
rep(1:4, c(2,1,2,1))	 # specify replicate number for each element individually
rep(1:4, each = 2, len = 4)    # first 4 only
rep(1:4, each = 2, len = 10)   # 8 integers plus two recycled 1's.
rep(1:4, each = 2, times = 3)  # length 24, 3 complete replications
```

### Challenge

Using the new functions you have learned we can now **create** a `data.frame` manually with the function `data.frame()`. This function takes in each column as an argument by giving it name, and the `c()` function concatenates the values into a vector.
Below we have re-created our metadata `data.frame`.

```{r, create-data}

create_data <- data.frame(genotype=rep(c("Wt", "KO"), each=3, len=12),
			  celltype=c(rep("typeA", 6), rep("typeB", 6)),
			  replicate=rep(seq(1,3), 4),
			  row.names=paste("sample", 1:12, sep=""))
```

## Dataframes

Our metadata frame has rows and columns (it has 2 dimensions), if we want to
extract some specific data from it, we need to specify the "coordinates" we want
from it. Row numbers come first, followed by column numbers.

```{r, purl=FALSE}
metadata[1, 1]   # first element in the first column of the data frame
metadata[1, 3]   # first element in the 6th column
metadata[1:3, ] # first three rows
metadata[3, ]    # the 3rd element for all columns
metadata[ ,3]    # the entire 3rd column
```

### Challenge

1. The function `nrow()` on a `data.frame` returns the number of rows. Use it,
   in conjuction with `seq()` to create a new `data.frame` called
   `surveys_by_10` that includes every 10th row of the survey data frame
   starting at row 10 (10, 20, 30, ...)

```{r, echo=FALSE, purl=TRUE}
### The function `nrow()` on a `data.frame` returns the number of
### rows. Use it, in conjuction with `seq()` to create a new
### `data.frame` called `surveys_by_10` that includes every 10th row
### of the survey data frame starting at row 10 (10, 20, 30, ...)
```

